A Domain-Specific Optimizing Compiler\\for Solving Partial Differential Equations

### 2014-03-05 paper/bare_conf.tex
General refactoring

We focus on Local Assembly (``assembly'', in the following), a fundamental step of a Finite Element Method that covers an important fraction of the overall computation run-time, typically in the range 30$\%$-60$\%$. During the assembly phase, the solution of the PDE is approximated by executing a suitable kernel over all elements in the discretized domain. A kernel's working set is usually small enough to fit the L1 cache; it might need L2 cache when high-order methods are employed to improve the accuracy of the solution. However, we do not consider the latter case. An assembly kernel is characterized by the presence of an affine, usually non-perfect loop nest, where individual loops are rather small (the trip count rarely exceeds 40, with a minimum value of 3, depending on the order of the method). With such small kernels, we focus on aspects like minimization of floating-point operations, register allocation and instruction-level parallelism, especially in the form of SIMD vectorisation. Our study is conducted in the context of Firedrake, a system for solving PDEs through the Finite Element Method based on the OP2 abstraction~\cite{firedrake}.

Optimization of assembly kernels is non-trivial. Polyhedral compilation could be used to improve performance; however, given the exceptionally small size of the kernels, effective register locality can be obtained only through an in-depth analysis of the data space, rather than by reasoning on the iteration space (our attempts to use PoCC's optimizations~\cite{PoCC}, especially unrolling and register tiling, did not improve performance over $icc$ compilation). BLAS routines could be employed as well, although fairly complicated code manipulation would be required, for a performance gain potentially far from the expectation. The major issue would concern automating identification and extraction of matrix-matrix multiplications from the code, which requires a non-trivial analysis of control- and data-flow. BLAS libraries are also known to perform distant from peak performance when the dimension of the matrices is small~\cite{nek5000}. In addition, register locality would be affected, since the same matrix could appear in a number of products. Hand-made BLAS-based implementations of the Helmholtz assembly kernel (illustrated later), varying the polynomial order from $p=1$ to $p=4$, has showed that the performance is worse than that achieved with our optimization strategy.

@@

Performance results show that Firedrake-generated code for non-trivial assembly kernels, if not adequately optimized, is sub-optimal. A cost-model-driven sequence of source-to-source code transformations, aimed at improving register allocation and register data locality, can result in performance improvements up to 33$\%$ over ``softly-optimized'' code (i.e. where only basic transformations are performed, such as auto-vectorizable loop-invariant code motion, padding and data alignment), and up to 77$\%$ over the original kernels. The contribution of this paper is therefore twofold:
\begin{itemize}
\item An optimisation strategy for a class of kernels widely-used in scientific applications, namely Local Assembly in the context of the Finite Element Method. Our approach exploits domain knowledge to go beyond the limits of both vendor (e.g. $icc$) and research (e.g. those based on the polyhedral model) compilers.
\item Automation of such code optimizations in Firedrake. This enables performing an in-depth performance evaluation of the proposed optimization strategy in real-world simulations.
\end{itemize}

@@

\section{Background}
%Here we discuss about the computational characteristics of computational science kernels. Briefly cite op2 kernels. Emphasis on Finite Element Assembly. Generalization and formalization of a Finite Element Assembly kernel using quadrature representation.
\begin{algorithm}[t]
\small
\caption{Local assembly code generated by Firedrake for a Helmholtz problem (2D mesh, Lagrange $p=1$ elements).}
\label{code:helmholtz}
\KwSty{void} helmholtz(double A[3][3], double **coords) $\lbrace$\\
~~// K, det = Compute Jacobian (coords) \\
~~\\
~~\KwSty{static const double} W3[3] = $\lbrace$...$\rbrace$\\
~~\KwSty{static const double} X$\_$D10[3][3] = $\lbrace\lbrace$...$\rbrace\rbrace$\\
~~\KwSty{static const double} X$\_$D01[3][3] = $\lbrace\lbrace$...$\rbrace\rbrace$\\
~~\\
~~\KwSty{for} (\KwSty{int} i = 0; i$<$3; i++) \\
~~~~\KwSty{for} (\KwSty{int} j = 0; j$<$3; j++) \\
~~~~~~\KwSty{for} (\KwSty{int} k = 0; k$<$3; k++) \\
~~~~~~~~A[j][k] += ((Y[i][k]*Y[i][j]+\\
~~~~~~~~~~~+((K[1]*X$\_$D10[i][k]+K[3]*X$\_$D01[i][k])*\\
~~~~~~~~~~~*(K[1]*X$\_$D10[i][j]+K[3]*X$\_$D01[i][j]))+\\
~~~~~~~~~~~+((K[0]*X$\_$D10[i][k]+K[2]*X$\_$D01[i][k])*\\
~~~~~~~~~~~*(K[0]*X$\_$D10[i][j]+K[2]*X$\_$D01[i][j])))*\\
~~~~~~~~~~~*det*W3[i]);\\
$\rbrace$
\end{algorithm}

\begin{algorithm}[t]
\small
\caption{Local assembly code generated by FFC for a Burgers problem (3D mesh, Lagrange $p=1$ elements).}
\label{code:burgers}
\KwSty{void} burgers(double A[12][12], double **c, double **w) $\lbrace$\\
~~// K, det = Compute Jacobian (c) \\
~~\\
~~\KwSty{static const double} W5[5] = $\lbrace$...$\rbrace$\\
~~\KwSty{static const double} X1$\_$D001[5][12] = $\lbrace\lbrace$...$\rbrace\rbrace$\\
~~\KwSty{static const double} X2$\_$D001[5][12] = $\lbrace\lbrace$...$\rbrace\rbrace$\\
~~//It follows 11 other basis functions definitions.\\
~~...\\
~~\\
~~\KwSty{for} (\KwSty{int} i = 0; i$<$5; i++) $\lbrace$\\
~~~~\KwSty{double} F0 = 0.0;\\
~~~~//It follows 10 other declarations (F1, F2,...)\\
~~~~...\\
~~~~\KwSty{for} (\KwSty{int} r = 0; r$<$12; r++) $\lbrace$\\
~~~~~~F0 += (w[r][0]*X1$\_$D100[ip][r]);\\
~~~~~~//It follows 10 analogous statements (F1, F2, ...)\\
~~~~$\rbrace$\\
~~~~...\\
~~~~\KwSty{for} (\KwSty{int} j = 0; j$<$12; j++) \\
~~~~~~\KwSty{for} (\KwSty{int} k = 0; k$<$12; k++) \\
~~~~~~~~A[j][k] += (..(K[5]*F9)+(K[8]*F10))*Y1[i][j])+\\
~~~~~~~~~+(((K[0]*X1$\_$D100[i][k])+(K[3]*X1$\_$D010[i][k])+\\
~~~~~~~~~+(K[6]*X1$\_$D001[i][k]))*Y2[i][j]))*F11)+\\
~~~~~~~~~+(..((K[2]*X2$\_$D100[i][k])+...+(K[8]*X2$\_$D001[i][k])*\\
~~~~~~~~~*(K[2]*X2$\_$D100[i][j])+...+(K[8]*X2$\_$D001[i][j]))..)+\\
~~~~~~~~~+ $<$roughly a hundred of sum/muls go here$>$)..)*\\
~~~~~~~~~*det*W5[i]);\\
~~$\rbrace$ \\
$\rbrace$
\end{algorithm}

Local assembly consists of evaluating so called element stiffness matrix and vector; in this work, we focus on computation of matrices, which is the costly part of the process. A stiffness matrix can be intuitively thought as an approximated representation of the PDE solution in a specific cell of the discretized domain. Numerical integration algorithms are widely-used in assembly codes to implement evaluation of stiffness matrices~\cite{quadrature1, quadrature2}.

@@

The complexity of Firedrake-generated kernels depends on the mathematical problem being solved. In simpler cases, the loop nest is perfect, it has short trip counts (in the range 3-15), and the computation reduces to a summation of a few products. An example is provided in Figure~\ref{code:helmholtz}, which shows an assembly kernel for a Helmholtz problem, using Lagrange basis functions on 2D elements with polynomial order $p=1$. The stiffness matrix in the code is called $A$. In other scenarios, for instance when solving a non-linear problem like Burgers (see Figure~\ref{code:burgers}), the number of arrays involved in the computation of $A$ can be much larger: in this case, 14 unique arrays are accessed, and the same array can be referenced multiple times within the expression. Also, constants evaluated in outer loops (called $F$ in the code), acting as scaling factors of arrays, may be required; trip counts can be larger (proportionally to the order of the method); arrays may be block-sparse. Note that in addition to a larger number of operations to compute the stiffness matrix, the Burgers case shows a register pressure higher than that in Helmholtz. Despite assembly kernels being problem-dependent, and so infinite is the space of codes that Firedrake can generate, it is still possible to identify common domain-specific traits, which can be exploited for effective code transformations and SIMD vectorization.

The class of kernels we are considering has, in particular, some peculiarities. 1) The computation of the Jacobian, which is the first step of the assembly, is independent of the loop nest; this is not true in general, since the unstructured mesh might use bended elements that would require the Jacobian be re-computed every $i$ iteration. 2) Memory accesses along the three loop dimensions are always 1-stride .3) The $j$ and $k$ loops are interchangeable, whereas permutation of $i$ is might be subjected to pre-computation of values and introduction of temporaries (e.g. the $F$ values in Burgers). 4) So called basis functions arrays (denoted by $X$, $Y$, ...) are constants, and most of the sub-expressions on the right hand side of the stiffness matrix computation depend on just two loops (either $i$-$j$ or $i$-$k$). In Section~\ref{sec:code-transf} we show how to exploit these observations to define a set of systematic, composable optimizations.

@@

\begin{algorithm}[t]
\small
\caption{Local assembly code generated by Firedrake for a Helmholtz problem (2D mesh, Lagrange $p=1$ elements). The padding, data alignment and $licm$ optimizations are applied. Data alignment and padding relate to an AVX machine. In this specific case, sub-expressions invariant to $j$ are identical to those invariant to $k$; in general, this could not be the case.}
\label{code:helmholtz-licm}
\KwSty{void} helmholtz(double A[3][4], double **coords) $\lbrace$\\
~~// K, det = Compute Jacobian (coords) \\
~~\\
~~\KwSty{static const double} W3[3] = $\lbrace$...$\rbrace$\\
~~\KwSty{static const double} X$\_$D10[3][4] = $\lbrace\lbrace$...$\rbrace\rbrace$\\
~~\KwSty{static const double} X$\_$D01[3][4] = $\lbrace\lbrace$...$\rbrace\rbrace$\\
~~\\
~~\KwSty{for} (\KwSty{int} i = 0; i$<$3; i++) $\lbrace$ \\
~~~~double LI$\_$0[4];\\
~~~~double LI$\_$1[4];\\
~~~~\KwSty{for} (\KwSty{int} r = 0; r$<$4; r++) $\lbrace$ \\
~~~~~~LI$\_$0[r] = ((K[1]*X$\_$D10[ip][r])+(K[3]*X$\_$D01[ip][r]));\\
~~~~~~LI$\_$1[r] = ((K[0]*X$\_$D10[ip][r])+(K[2]*X$\_$D01[ip][r]));\\
~~~~$\rbrace$\\
~~~~\KwSty{for} (\KwSty{int} j = 0; j$<$3; j++) \\
~~~~~~\KwSty{for} (\KwSty{int} k = 0; k$<$4; k++) \\
~~~~~~~~A[j][k] += (Y[i][k]*Y[i][j]+LI$\_$0[k]*LI$\_$0[j]+\\
~~~~~~~~~~~~~~~~~~~~+LI$\_$1[k]*LI$\_$1[j])*det*W3[i]);\\
~~$\rbrace$\\
$\rbrace$
\end{algorithm}

\begin{algorithm}[t]
\small
\caption{Local assembly code generated by Firedrake for a Helmholtz problem (2D mesh, Lagrange $p=1$ elements). The padding, data alignment, $licm$ and $split$ optimizations are applied. Data alignment and padding relate to an AVX machine. In this specific case, the split factor is 2.}
\label{code:helmholtz-split}
\KwSty{void} helmholtz(double A[3][4], double **coords) $\lbrace$\\
~~// K, det = Compute Jacobian (coords) \\
~~// Declaration of basis function matrices \\
~~\\
~~\KwSty{for} (\KwSty{int} i = 0; i$<$3; i++) $\lbrace$ \\
~~~~double LI$\_$0[4];\\
~~~~double LI$\_$1[4];\\
~~~~\KwSty{for} (\KwSty{int} r = 0; r$<$4; r++) $\lbrace$ \\
~~~~~~LI$\_$0[r] = ((K[1]*X$\_$D10[ip][r])+(K[3]*X$\_$D01[ip][r]));\\
~~~~~~LI$\_$1[r] = ((K[0]*X$\_$D10[ip][r])+(K[2]*X$\_$D01[ip][r]));\\
~~~~$\rbrace$\\
~~~~\KwSty{for} (\KwSty{int} j = 0; j$<$3; j++) \\
~~~~~~\KwSty{for} (\KwSty{int} k = 0; k$<$4; k++) \\
~~~~~~~~A[j][k] += (Y[i][k]*Y[i][j]+LI$\_$0[k]*LI$\_$0[j])*det*W3[i];\\
~~~~\KwSty{for} (\KwSty{int} j = 0; j$<$3; j++) \\
~~~~~~\KwSty{for} (\KwSty{int} k = 0; k$<$4; k++) \\
~~~~~~~~A[j][k] += LI$\_$1[k]*LI$\_$1[j]*det*W3[i];\\
~~$\rbrace$\\
$\rbrace$
\end{algorithm}

@@

% Autovectorizable LICM -> show code snippet?
\subsection{Auto-vectorizable Loop-invariant Code Motion}
From inspection of the codes in Figures~\ref{code:helmholtz} and~\ref{code:burgers}, it can be noticed that the computation of $A$ involves evaluating many sub-expressions that depend on two iteration variables only. Since symbols in most of these sub-expressions are read-only variables, there is ample space for loop-invariant code motion. Vendor compilers apply this technique, although not in the systematic way we need for our assembly kernels. We want to overcome two deficiencies that both $icc$ and $gcc$ have. First, these compilers can identify sub-expressions that are invariant with respect to the innermost loop only. This is an issue for sub-expressions depending on $i$-$k$, which are not automatically lifted. Second, the hoisted code is scalar, i.e. it is not subjected to auto-vectorization. We work around these limitations with source-level loop-invariant code motion. In particular, we pre-compute all values that an invariant sub-expression assumes along the fastest varying dimension. This is implemented by introducing a temporary array (per invariant sub-expression) and by adding a new loop to the nest. At the price of extra memory for storing temporaries, the gain is that lifted terms can be auto-vectorized, because part of an inner loop. Note that given the short trip counts of our loops, it is important to achieve auto-vectorization of all hoisted terms in order to minimize the percentage of scalar instructions, which could be otherwise significant.

@@

% Register allocation and register locality
\subsection{Optimizing Register Management}
One notable problem assembly kernels are exposed to concerns register allocation and register locality. The critical situation occurs when loop trip counts and accessed variables are such that the vector-registers pressure is high. Due to a kernel's working set fitting the L1 cache, it is remarkably important to optimize register management in order to maximize performance. Canonical optimizations, such as loop interchange, unroll, unroll-and-jam and register tiling are typically employed to deal with this problem. In the compiler we have developed, we support these optimizations either by means of explicit source code transformation (interchange, unroll-and-jam) or indirectly advising the compiler through standard pragmas (unroll). 
In this context, register tiling is a critical optimization. A way of obtaining register tiles consists of slicing the innermost loop into rectangular blocks of identical size (except the ``reminder'' block), and applying unrolling to individual blocks in order to expose ILP. This is automated by some vendor compilers~\cite{comdex}. Register tiles along both inner dimensions can be determined as well. Based on the observation that the evaluation of the stiffness matrix can be reduced to a ``summation of outer products'' along the $j$ and $k$ dimensions, we implement a domain-specific vector-register tiling strategy that leads to larger tiles, enhancing reuse. If we consider the code snippet in Figure~\ref{code:helmholtz-licm} (Helmholtz after loop-invariant code motion), we can notice that the computation of $A$ is abstractly expressible as
\begin{equation}
\label{outer-product}
A_{jk} = \sum_{\substack{
  x \in B' \subseteq B \\ 
  y \in B'' \subseteq B}} 
x_j\cdot y_k ~~~~~~ j,k = 0,...,4
\end{equation}
where $B$ is the set of all matrices (or temporaries) accessed in the kernel, whereas $B'$ and $B''$ are generic problem-dependent subsets. Without loss of generality, the presence of constants or other variables independent of both $j$ and $k$ can be momentarily neglected. Note that regardless of the specific input problem, the stiffness matrix computation is always reducible to this kind of form. Figure~\ref{fig:vect-by-vect} illustrates how we can evaluate 16 elements ($j,k=0,...,4$) of the stiffness matrix using just 2 vector registers (a 4$\times$4 tile), assuming $\vert B' \vert = \vert B'' \vert = 1$. Values in a register are shuffled each time a product is performed. Standard compiler auto-vectorization ($gcc$ and $icc$), instead, executes 4 broadcast operations (i.e. ``splat'' of a value over all of the register locations) along the outer dimension to perform the calculation, and would also need to keep between $f=1$ and $f=3$ extra registers to perform the same 16 evaluations if unroll-and-jam is found to be useful, with $f$ being the unroll-and-jam factor.

@@

The storage layout of $A$, however, is incorrect after the application of this ``outer-product vectorization'' ($op$-$vect$). We efficiently restore it with a sequence of vector shuffle following the pattern highlighted in Figure~\ref{fig:restore-layout}, executed once outside of the $ijk$ loop nest. The generated pseudo-code for the simple Helmholtz problem when using $op$-$vect$ is shown in Figure~\ref{code:helmholtz-opvect}.

@@

@@

>>>>

We focus on Local Assembly (``assembly'', in the following), a fundamental step of a Finite Element Method that covers an important fraction of the overall computation run-time, typically in the range 30$\%$-60$\%$. During the assembly phase, the solution of the PDE is approximated by executing a suitable kernel over all elements in the discretized domain. A kernel's working set is usually small enough to fit the L1 cache; it might need L2 cache when high-order methods are employed to improve the accuracy of the solution. However, we do not consider the latter case. An assembly kernel is characterized by the presence of an affine, usually non-perfect loop nest, where individual loops are rather small (the trip count rarely exceeds 40, with a minimum value of 3, depending on the order of the method). With such small kernels, we focus on aspects like minimization of floating-point operations, register allocation and instruction-level parallelism, especially in the form of SIMD vectorization. Our study is conducted in the context of Firedrake, a system for solving PDEs through the Finite Element Method based on the OP2 abstraction~\cite{firedrake}.

Optimization of assembly kernels is non-trivial. Polyhedral compilation could be used to improve performance; however, given the exceptionally small size of the kernels, effective register locality can be obtained only through an in-depth analysis of the data space, rather than by reasoning on the iteration space (our attempts to use PoCC's optimizations~\cite{PoCC}, especially unrolling and register tiling, did not improve performance over $icc$ compilation). BLAS routines could be employed as well, although fairly complicated code manipulation would be required, for a performance gain potentially far from the expectation. The major issue would concern automating identification and extraction of matrix-matrix multiplications from the code, which requires a non-trivial analysis of control- and data-flow. BLAS libraries are also known to perform distant from peak performance when the dimension of the matrices is small~\cite{nek5000}. In addition, register locality would be affected, since the same matrix could appear in a number of products. Hand-made BLAS-based implementations of the Helmholtz assembly kernel (illustrated later), varying the polynomial order from $p=1$ to $p=4$ and on a 3D mesh, have showed run-times worse than those achieved with our optimization strategy.

@@

Early experiments showed that Firedrake-generated code for non-trivial assembly kernels was sub-optimal. Our cost-model-driven sequence of source-to-source code transformations, aimed at improving SIMD vectorization and register data locality, can result in performance improvements up to 33$\%$ over ``softly-optimized'' code (i.e. where only basic transformations are performed, such as auto-vectorizable loop-invariant code motion, padding and data alignment), and up to 77$\%$ over original kernels. The contribution of this paper is therefore twofold:
\begin{itemize}
\item An optimisation strategy for a class of kernels widely-used in scientific applications, namely Local Assembly in the context of the Finite Element Method. Our approach exploits domain knowledge to go beyond the limits of both vendor (e.g. $icc$) and research (e.g. polyhedral) compilers.
\item Automation of such code optimizations in Firedrake. This enables performing an in-depth performance evaluation of the proposed optimization strategy in real-world simulations.
\end{itemize}

@@

\section{Background}
\label{sec:background}
%Here we discuss about the computational characteristics of computational science kernels. Briefly cite op2 kernels. Emphasis on Finite Element Assembly. Generalization and formalization of a Finite Element Assembly kernel using quadrature representation.
\begin{algorithm}[t]
\small
\caption{Local assembly code generated by Firedrake for a Helmholtz problem (2D mesh, Lagrange $p=1$ elements).}
\label{code:helmholtz}
\KwSty{void} helmholtz(double A[3][3], double **coords) $\lbrace$\\
~~// K, det = Compute Jacobian (coords) \\
~~\\
~~\KwSty{static const double} W3[3] = $\lbrace$...$\rbrace$\\
~~\KwSty{static const double} X$\_$D10[3][3] = $\lbrace\lbrace$...$\rbrace\rbrace$\\
~~\KwSty{static const double} X$\_$D01[3][3] = $\lbrace\lbrace$...$\rbrace\rbrace$\\
~~\\
~~\KwSty{for} (\KwSty{int} i = 0; i$<$3; i++) \\
~~~~\KwSty{for} (\KwSty{int} j = 0; j$<$3; j++) \\
~~~~~~\KwSty{for} (\KwSty{int} k = 0; k$<$3; k++) \\
~~~~~~~~A[j][k] += ((Y[i][k]*Y[i][j]+\\
~~~~~~~~~~~+((K1*X$\_$D10[i][k]+K3*X$\_$D01[i][k])*\\
~~~~~~~~~~~*(K1*X$\_$D10[i][j]+K3*X$\_$D01[i][j]))+\\
~~~~~~~~~~~+((K0*X$\_$D10[i][k]+K2*X$\_$D01[i][k])*\\
~~~~~~~~~~~*(K0*X$\_$D10[i][j]+K2*X$\_$D01[i][j])))*\\
~~~~~~~~~~~*det*W3[i]);\\
$\rbrace$
\end{algorithm}

\begin{algorithm}[t]
\small
\caption{Local assembly code generated by FFC for a Burgers problem (3D mesh, Lagrange $p=1$ elements).}
\label{code:burgers}
\KwSty{void} burgers(double A[12][12], double **c, double **w) $\lbrace$\\
~~// K, det = Compute Jacobian (c) \\
~~\\
~~\KwSty{static const double} W5[5] = $\lbrace$...$\rbrace$\\
~~\KwSty{static const double} X1$\_$D001[5][12] = $\lbrace\lbrace$...$\rbrace\rbrace$\\
~~\KwSty{static const double} X2$\_$D001[5][12] = $\lbrace\lbrace$...$\rbrace\rbrace$\\
~~//It follows 11 other basis functions definitions.\\
~~...\\
~~\\
~~\KwSty{for} (\KwSty{int} i = 0; i$<$5; i++) $\lbrace$\\
~~~~\KwSty{double} F0 = 0.0;\\
~~~~//It follows 10 other declarations (F1, F2,...)\\
~~~~...\\
~~~~\KwSty{for} (\KwSty{int} r = 0; r$<$12; r++) $\lbrace$\\
~~~~~~F0 += (w[r][0]*X1$\_$D100[i][r]);\\
~~~~~~//It follows 10 analogous statements (F1, F2, ...)\\
~~~~$\rbrace$\\
~~~~...\\
~~~~\KwSty{for} (\KwSty{int} j = 0; j$<$12; j++) \\
~~~~~~\KwSty{for} (\KwSty{int} k = 0; k$<$12; k++) \\
~~~~~~~~A[j][k] += (..(K5*F9)+(K8*F10))*Y1[i][j])+\\
~~~~~~~~~+(((K0*X1$\_$D100[i][k])+(K3*X1$\_$D010[i][k])+\\
~~~~~~~~~+(K6*X1$\_$D001[i][k]))*Y2[i][j]))*F11)+\\
~~~~~~~~~+(..((K2*X2$\_$D100[i][k])+...+(K8*X2$\_$D001[i][k])*\\
~~~~~~~~~*(K2*X2$\_$D100[i][j])+...+(K8*X2$\_$D001[i][j]))..)+\\
~~~~~~~~~+ $<$roughly a hundred of sum/muls go here$>$)..)*\\
~~~~~~~~~*det*W5[i]);\\
~~$\rbrace$ \\
$\rbrace$
\end{algorithm}

Local assembly consists of evaluating so called element stiffness matrix and element stiffness vector; in this work, we focus on computation of matrices, which is the costly part of the process. A stiffness matrix can be intuitively thought as an approximated representation of the PDE solution in a specific cell (or element) of the discretized domain. Numerical integration algorithms are widely-used in assembly codes to implement evaluation of stiffness matrices~\cite{quadrature1, quadrature2}.

@@

The complexity of Firedrake-generated kernels depends on the mathematical problem being solved. In simpler cases, the loop nest is perfect, it has short trip counts (in the range 3-15), and the computation reduces to a summation of a few products. An example is provided in Figure~\ref{code:helmholtz}, which shows an assembly kernel for a Helmholtz problem, using Lagrange basis functions on 2D elements with polynomial order $p=1$. The stiffness matrix in the code is called $A$. In other scenarios, for instance when solving a non-linear problem like Burgers (see Figure~\ref{code:burgers}), the number of arrays involved in the computation of $A$ can be much larger: in this case, 14 unique arrays are accessed, and the same array can be referenced multiple times within the expression. Also, constants evaluated in outer loops (called $F$ in the code), acting as scaling factors of arrays, may be required; trip counts can be larger (proportionally to the order of the method); arrays may be block-sparse. Note that in addition to a larger number of operations to compute the stiffness matrix, the Burgers case shows a register pressure higher than that in Helmholtz. Despite assembly kernels being problem-dependent, meaning that infinite is the space of codes that Firedrake can generate, it is still possible to identify common domain-specific traits, which can be exploited for effective code transformations and SIMD vectorization.

The class of kernels we are considering has, in particular, some peculiarities. 1) The computation of the Jacobian, which is the first step of the assembly, is independent of the loop nest. This is not true in general, since the unstructured mesh might use bended elements that would require the Jacobian be re-computed every $i$ iteration; 2) memory accesses along the three loop dimensions are always 1-stride; 3) the $j$ and $k$ loops are interchangeable, whereas permutation of $i$ might be subjected to pre-computation of values (e.g. the $F$ values in Burgers) and introduction of temporary arrays; 4) the $j$ and $k$ loops iterate over the same iteration space; 5) so called basis functions arrays (denoted by $X$, $Y$, ...) are constants, and most of the sub-expressions on the right hand side of the stiffness matrix computation depend on just two loops (either $i$-$j$ or $i$-$k$). In Section~\ref{sec:code-transf} we show how to exploit these observations to define a set of systematic, composable optimizations.

@@

\begin{algorithm}[t]
\small
\caption{Local assembly code generated by Firedrake for a Helmholtz problem (2D mesh, Lagrange $p=1$ elements). The padding, data alignment and $licm$ optimizations are applied. Data alignment and padding relate to an AVX machine. In this specific case, sub-expressions invariant to $j$ are identical to those invariant to $k$; in general, this could not be the case.}
\label{code:helmholtz-licm}
\KwSty{void} helmholtz(double A[3][4], double **coords) $\lbrace$\\
~~\KwSty{$\#$define} ALIGN $\_\_$attribute$\_\_$((aligned(32))) \\
~~// K, det = Compute Jacobian (coords) \\
~~\\
~~\KwSty{static const double} W3[3] ALIGN = $\lbrace$...$\rbrace$\\
~~\KwSty{static const double} X$\_$D10[3][4] ALIGN = $\lbrace\lbrace$...$\rbrace\rbrace$\\
~~\KwSty{static const double} X$\_$D01[3][4] ALIGN = $\lbrace\lbrace$...$\rbrace\rbrace$\\
~~\\
~~\KwSty{for} (\KwSty{int} i = 0; i$<$3; i++) $\lbrace$ \\
~~~~double LI$\_$0[4];\\
~~~~double LI$\_$1[4];\\
~~~~\KwSty{for} (\KwSty{int} r = 0; r$<$4; r++) $\lbrace$ \\
~~~~~~LI$\_$0[r] = ((K1*X$\_$D10[i][r])+(K3*X$\_$D01[i][r]));\\
~~~~~~LI$\_$1[r] = ((K0*X$\_$D10[i][r])+(K2*X$\_$D01[i][r]));\\
~~~~$\rbrace$\\
~~~~\KwSty{for} (\KwSty{int} j = 0; j$<$3; j++) \\
~~~~~~\KwSty{$\#$pragma vector aligned}\\
~~~~~~\KwSty{for} (\KwSty{int} k = 0; k$<$4; k++) \\
~~~~~~~~A[j][k] += (Y[i][k]*Y[i][j]+LI$\_$0[k]*LI$\_$0[j]+\\
~~~~~~~~~~~~~~~~~~~~+LI$\_$1[k]*LI$\_$1[j])*det*W3[i]);\\
~~$\rbrace$\\
$\rbrace$
\end{algorithm}

@@

% Autovectorizable LICM -> show code snippet?
\subsection{Auto-vectorizable Loop-invariant Code Motion}
From inspection of the codes in Figures~\ref{code:helmholtz} and~\ref{code:burgers}, it can be noticed that the computation of $A$ involves evaluating many sub-expressions that depend on two iteration variables only. Since symbols in most of these sub-expressions are read-only variables, there is ample space for loop-invariant code motion. Vendor compilers apply this technique, although not in the systematic way we need for our assembly kernels. We want to overcome two deficiencies that both $icc$ and $gcc$ have. First, these compilers can identify sub-expressions that are invariant with respect to the innermost loop only. This is an issue for sub-expressions depending on $i$-$k$, which are not automatically lifted. Second, the hoisted code is scalar, i.e. it is not subjected to auto-vectorization. We work around these limitations with source-level loop-invariant code motion. In particular, we pre-compute all values that an invariant sub-expression assumes along the fastest varying dimension. This is implemented by introducing a temporary array (per invariant sub-expression) and by adding a new loop to the nest. At the price of extra memory for storing temporaries, the gain is that lifted terms can be auto-vectorized, because part of an inner loop. Given the short trip counts of our loops, it is important to achieve auto-vectorization of hoisted terms in order to minimize the percentage of scalar instructions, which could be otherwise significant. It is also worth noting that, in some problems, invariant sub-expressions along $j$ are identical to those along $k$ (e.g. in Helmholtz). In these cases, we safely avoid redundant pre-computation since, as anticipated in Section~\ref{sec:background}, a property of our domain is that $j$ and $k$ loops share the same iteration space.

@@

% Register allocation and register locality
\subsection{Optimizing Register Management}
One notable problem assembly kernels are exposed to concerns register allocation and register locality. The critical situation occurs when loop trip counts and accessed variables are such that the vector-registers pressure is high. Due to a kernel's working set fitting the L1 cache, it is remarkably important to optimize register management. Canonical optimizations, such as loop interchange, unroll, unroll-and-jam and register tiling are typically employed to deal with this problem. In the compiler we have developed, we support these optimizations either by means of explicit source code transformation (interchange, unroll-and-jam) or indirectly advising the compiler through standard pragmas (unroll). 
In this context, register tiling is a critical optimization. A way of obtaining register tiles consists of slicing the innermost loop into rectangular blocks of identical size (except the ``reminder'' block), and applying unrolling to individual blocks in order to expose ILP. This is automated by some vendor compilers~\cite{comdex}. Register tiles along both inner dimensions can be determined as well. Based on the observation that the evaluation of the stiffness matrix can be reduced to a ``summation of outer products'' along the $j$ and $k$ dimensions, we implement a domain-specific vector-register tiling strategy that leads to larger tiles, enhancing reuse. If we consider the code snippet in Figure~\ref{code:helmholtz-licm} (Helmholtz after loop-invariant code motion), we can notice that the computation of $A$ is abstractly expressible as
\begin{equation}
\label{outer-product}
A_{jk} = \sum_{\substack{
  x \in B' \subseteq B \\ 
  y \in B'' \subseteq B}} 
x_j\cdot y_k ~~~~~~ j,k = 0,...,4
\end{equation}
where $B$ is the set of all matrices (or temporaries) accessed in the kernel, whereas $B'$ and $B''$ are generic problem-dependent subsets. Without loss of generality, the presence of constants or other variables independent of both $j$ and $k$ can be momentarily neglected. Note that regardless of the specific input problem, the stiffness matrix computation is always reducible to this kind of form. Figure~\ref{fig:vect-by-vect} illustrates how we can evaluate 16 elements ($j,k=0,...,4$) of the stiffness matrix using just 2 vector registers (a 4$\times$4 tile), assuming $\vert B' \vert = \vert B'' \vert = 1$. Values in a register are shuffled each time a product is performed. Standard compiler auto-vectorization ($gcc$ and $icc$), instead, executes 4 broadcast operations (i.e. ``splat'' of a value over all of the register locations) along the outer dimension to perform the calculation, and would also need to keep between $f=1$ and $f=3$ extra registers to perform the same 16 evaluations when unroll-and-jam is used, with $f$ being the unroll-and-jam factor.

@@

The storage layout of $A$, however, is incorrect after the application of this ``outer-product vectorization'' ($op$-$vect$). We efficiently restore it with a sequence of vector shuffles following the pattern highlighted in Figure~\ref{fig:restore-layout}, executed once outside of the $ijk$ loop nest. The generated pseudo-code for the simple Helmholtz problem when using $op$-$vect$ is shown in Figure~\ref{code:helmholtz-opvect}.

@@

\begin{algorithm}[t]
\small
\caption{Local assembly code generated by Firedrake for a Helmholtz problem (2D mesh, Lagrange $p=1$ elements). The padding, data alignment, $licm$ and $split$ optimizations are applied. Data alignment and padding relate to an AVX machine. In this specific case, the split factor is 2.}
\label{code:helmholtz-split}
\KwSty{void} helmholtz(double A[3][4], double **coords) $\lbrace$\\
~~\KwSty{$\#$define} ALIGN $\_\_$attribute$\_\_$((aligned(32))) \\
~~// K, det = Compute Jacobian (coords) \\
~~// Declaration of basis function matrices \\
~~\\
~~\KwSty{for} (\KwSty{int} i = 0; i$<$3; i++) $\lbrace$ \\
~~~~double LI$\_$0[4];\\
~~~~double LI$\_$1[4];\\
~~~~\KwSty{for} (\KwSty{int} r = 0; r$<$4; r++) $\lbrace$ \\
~~~~~~LI$\_$0[r] = ((K1*X$\_$D10[i][r])+(K3*X$\_$D01[i][r]));\\
~~~~~~LI$\_$1[r] = ((K0*X$\_$D10[i][r])+(K2*X$\_$D01[i][r]));\\
~~~~$\rbrace$\\
~~~~\KwSty{for} (\KwSty{int} j = 0; j$<$3; j++) \\
~~~~~~\KwSty{$\#$pragma vector aligned}\\
~~~~~~\KwSty{for} (\KwSty{int} k = 0; k$<$4; k++) \\
~~~~~~~~A[j][k] += (Y[i][k]*Y[i][j]+LI$\_$0[k]*LI$\_$0[j])*det*W3[i];\\
~~~~\KwSty{for} (\KwSty{int} j = 0; j$<$3; j++) \\
~~~~~~\KwSty{$\#$pragma vector aligned}\\
~~~~~~\KwSty{for} (\KwSty{int} k = 0; k$<$4; k++) \\
~~~~~~~~A[j][k] += LI$\_$1[k]*LI$\_$1[j]*det*W3[i];\\
~~$\rbrace$\\
$\rbrace$
\end{algorithm}

@@

Mention heuristic used by our compiler for splitting.. TODO
